---
title: "DHIS2 Research Project"
author: ""
date: "24th Mar, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
  pdf_document:
    toc: true
  word_document: 
    toc: true
    fig_width: 7
    fig_height: 5
editor_options:
  chunk_output_type: console
---

<style>
h1.title {text-align: center;margin-bottom: 30px;}
h4.author {text-align: center;margin-top: 10px;margin-bottom: 20px;}
h4.date {text-align: center;margin-top: 10px;margin-bottom: 40px;}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)
```


<br> 


############################################################
#### Population data Interpolation at exponential growth ###
############################################################


```{r}
# Annual population data
annual_pop = data.frame(year = 2015:2025,
                        population = c(37531446, 38799152,
                                       40177388, 41565831,
                                       42999637, 44457152,
                                       45910930, 47312719,
                                       48656601, 50015092,
                                       51384894))

# Interpolation at exponential growth
monthly_exponential = do.call(
  rbind,
  lapply(seq_len(nrow(annual_pop) - 1), function(i) {
    start_pop  = annual_pop$population[i]
    end_pop    = annual_pop$population[i+1]
    curr_year = annual_pop$year[i+1]
    
r_monthly = (end_pop / start_pop)^(1/12) - 1
    
data.frame(year  = rep(curr_year,12),
           month = 1:12,
           population = start_pop * (1 + r_monthly)^(1:12))}))

head(monthly_exponential, 12)

```


```{r}
monthly_exponential$year_month = as.Date(
  paste(monthly_exponential$year,
        monthly_exponential$month,
        "01", sep = "-"))

monthly_exponential$period = format(
  monthly_exponential$year_month, "%b-%y")

library(ggplot2)
pop = ggplot(monthly_exponential,
       aes(x = year_month, y = population/10^6)) +
  geom_line(linewidth = 1) +
  geom_point( data = subset(monthly_exponential, month == 12),
              size = 2) +
  labs(x = "Period", y = "Population in millions") +
  theme_bw()
pop

```


<br> 

############################################################
##################### Trends analysis ######################
############################################################


```{r}

library(tidyverse)
library(lubridate)
library(MASS)        # Negative binomial
library(forecast)    # ARIMA, STL
library(tsoutliers)  # Outlier detection
library(broom)
library(AER)         # Overdispersion test
library(ggplot2)
library(ggplotify)

set.seed(100)

stroke_data1 = read_csv("G:/Dataset1.csv")

stroke_data2 = read_csv("G:/Dataset2.csv")

stroke_data = rbind(stroke_data1, stroke_data2)
View(stroke_data)

stroke_data = merge(
  stroke_data,
  monthly_exponential[, c("period", "population")],
  by = "period", all.x = TRUE)

stroke_data$date = as.Date(
  paste0("01-", stroke_data$period),
  format = "%d-%b-%y")

stroke_data = stroke_data[order(stroke_data$date), ]

stroke_data = stroke_data %>%
  mutate(time = 1:n(),
         month = factor(month(my(period)), levels=1:12),
         year = year(date))

```


### Dealing with outliers

```{r}
# Convert to time series
y = stroke_data$`stroke accidents`
y_ts = ts(y, start=c(2016,1), frequency=12)

# Detect and replace outliers using ARIMA
outliers = tso(y_ts, types="AO", tsmethod = "auto.arima")
y_ts_clean = outliers$yadj
stroke_data$stroke_accidents_clean = round(as.numeric(y_ts_clean),0)



y = stroke_data$`stroke cases`
y_ts = ts(y, start=c(2020,1), frequency=12)

outliers = tso(y_ts, types="AO", tsmethod = "auto.arima")
y_ts_clean = outliers$yadj
stroke_data$stroke_cases_clean = round(as.numeric(y_ts_clean),0)



y = stroke_data$`stroke deaths`
y_ts = ts(y, start=c(2020,1), frequency=12)

outliers = tso(y_ts, types="AO", tsmethod = "auto.arima")
y_ts_clean = outliers$yadj
stroke_data$stroke_deaths_clean = round(as.numeric(y_ts_clean),0)

```


### Summarizing data

```{r}
library(dplyr)
library(lubridate)
class(stroke_data$date)

annual = stroke_data %>% mutate(year = year(date)) %>%
  group_by(year) %>%
  summarise(suspected = sum(stroke_accidents_clean),
            confirmed = sum(stroke_cases_clean),
            deaths = sum(stroke_deaths_clean),
            population = mean(population)) %>%
  mutate(incidence_rate = round(confirmed/population*100000,3),
         mortality_rate = round(deaths/population*100000,3),
         CFR = round(deaths/confirmed,3))
annual

write.csv(annual, "G:/My Drive/PhD/Manuscripts/DHIS2/Datasets/stroke_annual_summary.csv")


```


### Monthly stroke rates trends

```{r}

library(ggplot2)
library(dplyr)
library(scales)

stroke_monthly = stroke_data %>% mutate(year = year(date),
         incidence_rate = stroke_cases_clean/population*100000,
         mortality_rate = stroke_deaths_clean/population*100000,
         CFR = stroke_deaths_clean/stroke_cases_clean)


stroke_monthly = stroke_monthly %>%
  mutate(date = as.Date(paste(year, month, "01", sep = "-")))

scale_factor = 3.5
stroke_monthly = stroke_monthly %>%
  mutate(CFR_scaled = CFR * scale_factor)
library(tidyr)
plot_data = stroke_monthly %>%
  dplyr::select(date, incidence_rate, mortality_rate, CFR_scaled) %>%
  pivot_longer(cols = c(incidence_rate, mortality_rate, CFR_scaled),
               names_to = "metric",
               values_to = "value")


DesB = ggplot(plot_data, aes(x = date, y = value, color = metric)) +
  geom_point(aes(shape = metric)) +
  geom_line(size = 0.8) +
  scale_y_continuous(name = "Rate per 100,000 population",
                     sec.axis = sec_axis(~ ./scale_factor,
                                      name = "Case fatality rate")) +
  scale_x_date(date_labels = "%b %Y") +
  theme_minimal() +
  theme(legend.position = c(0.15, 0.85),
        legend.key.size = unit(0.4, "cm"),
        legend.margin = margin(t = -5),
        legend.box.margin = margin(t = -10)) +
  scale_color_manual(values = c("CFR_scaled" = "coral",
                                "incidence_rate" = "skyblue",
                                "mortality_rate" = "red2"),
                     labels = c("Case fatality rate",
                                "Incidence rate",
                                "Mortality rate"),
                     name = "") +
  labs(title = "B", x = "Period") + 
  scale_shape_manual(values = c("CFR_scaled" = 25,
                                "incidence_rate" = 15,
                                "mortality_rate" = 13),
                     labels = c("Case fatality rate",
                                "Incidence rate",
                                "Mortality rate"),
                     name = "")

```


### Plot Suspected vs Confirmed Cases (Monthly)

```{r}
library(tidyr)

stroke_long = stroke_monthly %>%
  dplyr::select(date, stroke_accidents_clean,
                stroke_cases_clean) %>%
  pivot_longer(cols = c(stroke_accidents_clean,
                        stroke_cases_clean),
               names_to = "type",
               values_to = "cases")

DesA = ggplot(stroke_long, aes(x = date, y = cases, color = type)) +
  geom_point(aes(shape = type)) +
  geom_line(linewidth = 0.8) +
  labs(title = "A", x = "Period", y = "Number of Cases",
       color = "Case Type") +
  scale_color_manual(name = "",
    values = c("stroke_accidents_clean" = "steelblue",
               "stroke_cases_clean" = "coral2"),
    labels = c("Suspected cases", "Confirmed cases")) +
  theme_minimal() + 
  theme(legend.position = c(0.15, 0.93),
        legend.key.size = unit(0.4, "cm"),
        legend.margin = margin(t = -5),
        legend.box.margin = margin(t = -10)) +
  ylim(0,2700) + 
  scale_shape_manual(values = c("stroke_accidents_clean" = 25,
                                "stroke_cases_clean" = 13),
                     labels = c("Suspected cases",
                                "Confirmed cases"),
                     name = "")

```

### Descriptive plots merged

```{r}
library(ggpubr)
ggarrange(DesA, DesB, ncol=1, nrow=2)
```


### Stroke cases

```{r}

# Poisson model
pois_model = glm(stroke_cases_clean ~ time +
                   offset(log(population)),
                  family=poisson(link="log"), 
                 data=stroke_data)

dispersiontest(pois_model) # dispersion test

# Negative Binomial model
nb_model = glm.nb(stroke_cases_clean ~ time +
                    offset(log(population)),
                  data=stroke_data)


data.frame(Model = c("Poisson model",
                     "Negative Binomial model"),
           AIC = c(AIC(pois_model),AIC(nb_model)))


# Bayesian Negative Binomial model
library(brms)
bayes_nb = brm(stroke_cases_clean ~ time +
                 offset(log(population)),
               family = negbinomial(),
               data = stroke_data,
               prior = c(prior(normal(0,10),
                               class = "Intercept"),
                         prior(normal(0,1), class = "b")),
               chains = 4,
               iter = 1000)
summary(bayes_nb)
pp_check(bayes_nb)

# Bayesian Structural time series
library(bsts)
stroke_data$log_acc = log(stroke_data$stroke_cases_clean + 1)
ss <- list()
ss <- AddLocalLinearTrend(ss, stroke_data$log_acc)
ss <- AddSeasonal(ss, stroke_data$log_acc, nseasons = 12)

model_bsts <- bsts(log_acc ~ time, state.specification = ss,
                   data = stroke_data, niter = 1000)
plot(model_bsts)
summary(model_bsts)

# Generalized Additive model (NB)
library(mgcv)
stroke_data$time <- as.numeric(stroke_data$time)
stroke_data$month <- as.numeric(stroke_data$month)
gam_nb = gam(stroke_cases_clean ~ s(time, k = 20) +
                s(month, bs = "cc", k = 12) +
                offset(log(population)),
              family = nb(), data = stroke_data, method = "REML")
summary(gam_nb)
plot(gam_nb, pages = 1)
gam.check(gam_nb)

# Generalized Additive model LSS (NB)
library(gamlss)
gam_lss = gamlss(stroke_cases_clean ~ pb(time) + pb(month) +
                   offset(log(population)), 
                 sigma.formula = ~ pb(time),
                 family = NBII(), data = stroke_data)
plot(gam_lss)

# Compare predictive performance
train <- stroke_data[1:96, ]
test  <- stroke_data[97:120, ]
# 
# # On training set
# pred_nb <- predict(nb_model, newdata = train, type = "response")
# pred_bayes <- posterior_predict(bayes_nb, newdata = train)
# pred_bayes_mean <- colMeans(pred_bayes)
# pred_bsts <- predict(model_bsts, newdata = train, type = "response")
# pred_bsts_mean <- pred_bsts$mean
# pred_gam <- predict(gam_nb, newdata = train, type = "response")
# pred_gamlss <- predict(gam_lss, newdata = train, type = "response")
# 
# rmse_nb <- sqrt(mean((train$stroke_accidents_clean - pred_nb)^2))
# rmse_bayes <- sqrt(mean((train$stroke_accidents_clean - pred_bayes_mean)^2))
# rmse_bsts <- sqrt(mean((train$stroke_accidents_clean - pred_bsts_mean)^2))
# rmse_gam <- sqrt(mean((train$stroke_accidents_clean - pred_gam)^2))
# rmse_gamlss <- sqrt(mean((train$stroke_accidents_clean - pred_gamlss)^2))
# 
# mae_nb <- mean(abs(train$stroke_accidents_clean - pred_nb))
# mae_bayes <- mean(abs(train$stroke_accidents_clean - pred_bayes_mean))
# mae_bsts <- mean(abs(train$stroke_accidents_clean - pred_bsts_mean))
# mae_gam <- mean(abs(train$stroke_accidents_clean - pred_gam))
# mae_gamlss <- mean(abs(train$stroke_accidents_clean - pred_gamlss))



# On test set
pred_nb <- predict(nb_model, newdata = test, type = "response")
pred_bayes <- posterior_predict(bayes_nb, newdata = test)
pred_bayes_mean <- colMeans(pred_bayes)
pred_bsts <- predict(model_bsts, newdata = test, type = "response")
pred_bsts_mean <- pred_bsts$mean
pred_gam <- predict(gam_nb, newdata = test, type = "response")
pred_gamlss <- predict(gam_lss, newdata = test, type = "response")

rmse_nb <- sqrt(mean((test$stroke_accidents_clean - pred_nb)^2))
rmse_bayes <- sqrt(mean((test$stroke_accidents_clean - pred_bayes_mean)^2))
rmse_bsts <- sqrt(mean((test$stroke_accidents_clean - pred_bsts_mean)^2))
rmse_gam <- sqrt(mean((test$stroke_accidents_clean - pred_gam)^2))
rmse_gamlss <- sqrt(mean((test$stroke_accidents_clean - pred_gamlss)^2))


mae_nb <- mean(abs(test$stroke_accidents_clean - pred_nb))
mae_bayes <- mean(abs(test$stroke_accidents_clean - pred_bayes_mean))
mae_bsts <- mean(abs(test$stroke_accidents_clean - pred_bsts_mean))
mae_gam <- mean(abs(test$stroke_accidents_clean - pred_gam))
mae_gamlss <- mean(abs(test$stroke_accidents_clean - pred_gamlss))


metrics = data.frame(Model = c("NB", "BNB", "BSTS",
                               "GAM", "GAMLSS"),
                     RMSE = c(rmse_nb, rmse_bayes,
                              rmse_bsts, rmse_gam, rmse_gamlss),
                     MAE = c(mae_nb, mae_bayes, mae_bsts,
                             mae_gam, mae_gamlss))

metrics

# Extract estimates
summary(bayes_nb)
post <- posterior_summary(bayes_nb)

irr <- exp(post[, "Estimate"])
lower <- exp(post[, "Q2.5"])
upper <- exp(post[, "Q97.5"])

results <- data.frame(IRR = irr, CI_lower = lower, CI_upper = upper)
results[grep("^b_", rownames(results)), ]


# Residual inspection
res = residuals(bayes_nb)[, "Estimate"]
Box.test(res, lag = 12, type = "Ljung-Box")

### Decomposition: Population Growth vs Epidemiological Change

stroke_inc = stroke_data %>%
  mutate(fitted_cases = predict(bayes_nb, newdata =.,
                                type = "response"),
         fitted_rate = (fitted_cases[, "Estimate"] / population) * 100000)

annual_cases = stroke_inc %>% group_by(year) %>%
  summarise(population = mean(population),
            cases = sum(stroke_cases_clean),
          cases_model = sum(fitted_cases[, "Estimate"]),
          rate_model = sum(fitted_cases[, "Estimate"]) / mean(population) * 100000) %>%
  arrange(year)

annual_cases = annual_cases %>%
  mutate(pop_effect = (population - lag(population))*
      ((rate_model + lag(rate_model))/2) / 100000,
      
      rate_effect = (rate_model - lag(rate_model))*
      ((population + lag(population))/2) / 100000,
      
      case_change = cases_model - lag(cases_model))

# Rates per 100,000 population
annual_cases = annual_cases %>%
  mutate(pop_effect_rate = (pop_effect / population)* 100000,
         rate_effect_rate = (rate_effect / population)* 100000)

write.csv(annual_cases, "G:/My Drive/PhD/Manuscripts/DHIS2/Datasets/Incidence Decomposition.csv")


# Plot
annual_long = annual_cases %>%
  dplyr::select(year, pop_effect_rate, rate_effect_rate) %>%
  pivot_longer(-year, names_to="component", values_to="rate")

EpiA = ggplot(annual_long, aes(x=year, y=rate,
                               color=component)) +
  geom_line(linewidth=0.8) +
  geom_point(aes(shape = component), size=2) +
  geom_text(aes(label = round(rate, 3)), size = 3,
            vjust = -0.6, show.legend = FALSE) +
  labs(x="Period", y="Incidence rate per 100,000 \n population", color="Component") +
  scale_color_manual(name = "",
                     values=c("pop_effect_rate"="steelblue",
                              "rate_effect_rate"="tomato"),
                     labels = c("Population effect",
                                "Epidemiological effect")) +
  scale_shape_manual(name = "",
                     values=c("pop_effect_rate"=21,
                              "rate_effect_rate"=16),
                     labels = c("Population effect",
                                "Epidemiological effect")) +
  theme_minimal() + ylim(0,1.2) + 
  theme(legend.position = c(0.15, 0.9),
        legend.key.size = unit(0.4, "cm")) +
  scale_x_continuous(breaks = seq(min(annual_long$year),
                                max(annual_long$year),
                                by = 1))
```




### Stroke deaths

```{r}

# Poisson model
pois_model = glm(stroke_deaths_clean ~ time +
                   offset(log(population)),
                  family=poisson(link="log"), 
                 data=stroke_data)

dispersiontest(pois_model) # dispersion test

# Negative Binomial model
nb_model = glm.nb(stroke_deaths_clean ~ time +
                    offset(log(population)),
                  data=stroke_data)

data.frame(Model = c("Poisson model",
                     "Negative Binomial model"),
           AIC = c(AIC(pois_model), AIC(nb_model)))

# Bayesian Negative Binomial model
library(brms)
bayes_nb = brm(stroke_deaths_clean ~ time +
                 offset(log(population)),
               family = negbinomial(),
               data = stroke_data,
               prior = c(prior(normal(0,10),
                               class = "Intercept"),
                         prior(normal(0,1), class = "b")),
               chains = 4,
               iter = 1000)
summary(bayes_nb)
pp_check(bayes_nb)

# Extract estimates
summary(bayes_nb)
post <- posterior_summary(bayes_nb)

irr <- exp(post[, "Estimate"])
lower <- exp(post[, "Q2.5"])
upper <- exp(post[, "Q97.5"])

results <- data.frame(IRR = irr, CI_lower = lower, CI_upper = upper)
results[grep("^b_", rownames(results)), ]

# Residual inspection
res = residuals(bayes_nb)[, "Estimate"]
Box.test(res, lag = 12, type = "Ljung-Box")



### Decomposition: Population Growth vs Epidemiological Change

stroke_deaths = stroke_data %>%
  mutate(fitted_deaths = predict(bayes_nb, newdata =.,
                                type = "response"),
         fitted_rate = (fitted_deaths[, "Estimate"] / population) * 100000)

annual_deaths = stroke_deaths %>% group_by(year) %>%
  summarise(population = mean(population),
            deaths = sum(stroke_deaths_clean),
            deaths_model = sum(fitted_deaths[, "Estimate"]),
            rate_model = sum(fitted_deaths[, "Estimate"]) / mean(population) * 100000) %>%
  arrange(year)

annual_deaths = annual_deaths %>%
  mutate(pop_effect = (population - lag(population))*
      ((rate_model + lag(rate_model))/2) / 100000,
      
      rate_effect = (rate_model - lag(rate_model))*
      ((population + lag(population))/2) / 100000,
      
      deaths_change = deaths_model - lag(deaths_model))

# Rates per 100,000 population
annual_deaths = annual_deaths %>%
  mutate(pop_effect_rate = (pop_effect / population)* 100000,
         rate_effect_rate = (rate_effect / population)* 100000)

write.csv(annual_deaths, "G:/My Drive/PhD/Manuscripts/DHIS2/Datasets/mortality Decomposition.csv")

# Plot
annual_long = annual_deaths %>%
  dplyr::select(year, pop_effect_rate, rate_effect_rate) %>%
  pivot_longer(-year, names_to="component", values_to="rate")

EpiB = ggplot(annual_long, aes(x=year, y=rate,
                               color=component)) +
  geom_line(linewidth=0.8) + 
    geom_point(aes(shape = component), size=2) +
  geom_text(aes(label = round(rate, 3)), size = 3,
            vjust = -0.6, show.legend = FALSE) +
  labs(x="Period", y="Mortality rate per 100,000 \n population", color="Component") +
  scale_color_manual(name = "",
                     values=c("pop_effect_rate"="steelblue",
                              "rate_effect_rate"="tomato"),
                     labels = c("Population effect",
                                "Epidemiological effect")) +
  scale_shape_manual(name = "",
                     values=c("pop_effect_rate"=21,
                              "rate_effect_rate"=16),
                     labels = c("Population effect",
                                "Epidemiological effect")) +
  theme_minimal() + ylim(0, 0.15) + 
  theme(legend.position = c(0.15, 0.9),
        legend.key.size = unit(0.4, "cm")) +
  scale_x_continuous(breaks = seq(min(annual_long$year),
                                max(annual_long$year),
                                by = 1))

```

### Model-Based Decomposition plots merged

```{r}

library(ggpubr)
ggarrange(EpiA, EpiB, ncol=1, nrow=2)

```




### Case Fatality Trend

```{r}
# cfr_model = glm(
#   cbind(stroke_deaths_clean, stroke_cases_clean - stroke_deaths_clean) ~
#     time, family = binomial,data = stroke_data)
# 
# summary(cfr_model)
# 
# beta = coef(cfr_model)["time"]
# MPC = (exp(beta) - 1) * 100
# MPC

```




####################################################################
############################# Forecasting ##########################
####################################################################



```{r, include=FALSE}
library(dplyr)
library(readr)
library(janitor)
library(tidyverse)
library(ggplot2)
library(tibble)
library(tsibble)
library(forecast)
library(ggtime)
library(feasts)
library(fable)
library(fabletools)
library(seasonal)
library(ggplot2)

```



```{r}
stroke_data = stroke_data %>% 
  dplyr::select(-c(`stroke accidents`,
                   `stroke cases`,
                   `stroke deaths`))

stroke_data = stroke_data %>% 
  mutate(period = as.Date(paste0("01-", period),
                          format = "%d-%b-%y"),
         Month = yearmonth(period))

# Convert data to tibble
stroke_data = stroke_data %>%
  as_tsibble(index = Month)
```



####################
### Stroke cases
####################

```{r}

stroke_data %>%
  ggplot(aes(x = date, y = stroke_cases_clean)) +
  geom_line(linewidth = 0.8, color = "navyblue") +
  geom_point(size = 1) +
  scale_x_date(date_labels = "%b-%Y", date_breaks = "3 month") +
  labs(x = "Month", y = "Stroke Accidents") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Assess stationarity and seasonality

```{r}

stroke_data %>%
  gg_tsdisplay((stroke_cases_clean), plot_type='partial', lag=36) # These now appear to be stationary

# Unitroot kpss test. Ho: Data are stationary and non seasonal
stroke_data %>% 
  features(stroke_cases_clean, unitroot_kpss) # No evidence of H0

# Another way to determine if you need to differentiate based on STL decomposition. Ho: if the Seasonal strength; Fs > 0.64, take the 1st difference
stroke_data %>% features(stroke_cases_clean, feat_stl)

##### Seasonal plot
s3 = stroke_data %>%
  fill_gaps() %>%
  gg_season(stroke_cases_clean, labels = "none") +
  labs(y = "Confirmed stroke cases") +
  theme_classic()

### Seasonal Subseries plot
s4 = stroke_data %>%
  gg_subseries(stroke_cases_clean) +
  labs(y = "Confirmed stroke cases")


```

# Forecasting

```{r}

train = stroke_data %>%
  filter(year(Month) <= 2024)

acc_models = train %>%
  model(NAIVE = NAIVE(sqrt(stroke_cases_clean)),
        ETS = ETS(sqrt(stroke_cases_clean) ~ error("A") + trend("A") + season("A")),
        ARIMA = ARIMA(sqrt(stroke_cases_clean)))

acc_models %>% pivot_longer(everything(), names_to = "Model name",
                     values_to = "Orders")

data_eval = accuracy(acc_models) %>% arrange(RMSE) %>%
  dplyr::select(-c(.type, ACF1, ME, MPE))

# Evaluation Heat Map
df_long = data_eval %>%
  pivot_longer(cols = - .model, names_to = "Metric", values_to = "Value")

cases_metrics = ggplot(df_long, aes(x = Metric, y = .model, fill = Value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Value, 3)), size = 4) +
  scale_fill_gradient(low = "#f0f9e8", high = "#08589e") +
  labs(title = "Stroke cases",
       x = "Metrics", y = "Model", fill = "Value") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(hjust = 1),
    panel.grid = element_blank()) + guides(fill = "none")
cases_metrics

acc_fc = acc_models %>% forecast(h = "4 years") 
  
acc_fc %>% autoplot(stroke_data, level=NULL) +
  theme_classic()

acc_fc %>% accuracy(stroke_data) %>% 
  dplyr::select(-c(.type, ME, MPE, ACF1)) %>%
  arrange(RMSE)

# Best fit model: need to generate the 95% PI)
acc_futures = acc_models %>%
  generate(h = "4 years", times = 100) %>%
  as_tibble() %>% group_by(Month, .model) %>%
  summarise(dist = distributional::dist_sample(list(.sim))) %>%
  ungroup() %>%
  as_fable(index = Month, key = .model,
           distribution = dist,
           response = "stroke_cases_clean")

# Check the accuracy of the 95% predicted intervals (lower is better)
acc_futures %>% accuracy(stroke_data, 
                         measures = interval_accuracy_measures,
                         level = 95) %>% arrange(winkler)

# Extract Dec-2028 forecast at 80% from ETS
ets_dec_2028 = acc_fc %>%
  filter(.model == "ETS",
         Month == yearmonth("2028 Dec"))  %>%
  hilo(level=80) %>%
  unpack_hilo(`80%`) %>%
  rename(mean = .mean, LCI = `80%_lower`, UCI = `80%_upper`) %>%
  mutate(mean = round(mean),
         LCI = round(LCI),
         UCI = round(UCI)) 

cases_forecast = acc_futures %>% filter(.model == "ETS") %>%
  autoplot(stroke_data) + theme_bw() +
  labs(x = "Period",
       y = "Stroke cases") +
    theme(legend.position = c(0.02, 0.96),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", 
                                     colour = "black")) +
  geom_vline(xintercept = yearmonth("2028 Dec"),
             linetype = "dashed", colour = "red3") +
  geom_hline(yintercept = ets_dec_2028$mean,
             linetype = "dashed", colour = "red3") +
  geom_point(data = ets_dec_2028,
             aes(x = Month, y = mean),
             size = 2, colour = "red3") +
  geom_text(data = ets_dec_2028,
            aes(x = Month, y = mean,
                label = paste0("(2028 Dec, ", mean, ")",
                "\n", "80% CI (",LCI, ", ", UCI,")")),
            vjust = -0.3, hjust = 1.05, size = 3, 
            fontface = "bold", colour = "red3")
cases_forecast

```







#################
### Stroke deaths
#################

```{r}

stroke_data %>%
  ggplot(aes(x = date, y = stroke_deaths_clean)) +
  geom_line(linewidth = 0.8, color = "navyblue") +
  geom_point(size = 1) +
  scale_x_date(date_labels = "%b-%Y", date_breaks = "3 month") +
  labs(x = "Month", y = "Stroke Accidents") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Assess stationarity and seasonality

```{r}

stroke_data %>%
  gg_tsdisplay((stroke_deaths_clean), plot_type='partial', lag=36) # These now appear to be stationary

# Unitroot kpss test. Ho: Data are stationary and non seasonal
stroke_data %>% 
  features(stroke_deaths_clean, unitroot_kpss) # No evidence of H0

# Another way to determine if you need to differentiate based on STL decomposition. Ho: if the Seasonal strength; Fs > 0.64, take the 1st difference
stroke_data %>% features(stroke_deaths_clean, feat_stl)

##### Seasonal plot
s5 = stroke_data %>%
  fill_gaps() %>%
  gg_season(stroke_deaths_clean, labels = "none") +
  labs(y = "Stroke deaths") +
  theme_classic()

### Seasonal Subseries plot
s6 = stroke_data %>%
  gg_subseries(stroke_deaths_clean) +
  labs(y = "Stroke deaths")


```

# Forecasting

```{r}

train = stroke_data %>%
  filter(year(Month) <= 2024)

acc_models = train %>%
  model(NAIVE = NAIVE(sqrt(stroke_deaths_clean)),
        ETS = ETS(sqrt(stroke_deaths_clean) ~ error("A") + trend("A") + season("A")),
        ARIMA = ARIMA(sqrt(stroke_deaths_clean)))

acc_models %>% pivot_longer(everything(), names_to = "Model name",
                     values_to = "Orders")

data_eval = accuracy(acc_models) %>% arrange(RMSE) %>%
  dplyr::select(-c(.type, ACF1, ME, MPE))

# Evaluation Heat Map
df_long = data_eval %>%
  pivot_longer(cols = - .model, names_to = "Metric", values_to = "Value")

deaths_metrics = ggplot(df_long, aes(x = Metric, y = .model, fill = Value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Value, 3)), size = 4) +
  scale_fill_gradient(low = "#f0f9e8", high = "#08589e") +
  labs(title = "Stroke deaths",
       x = "Metrics", y = "Model", fill = "Value") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(hjust = 1),
    panel.grid = element_blank()) + guides(fill = "none")
deaths_metrics

acc_fc = acc_models %>% forecast(h = "4 years") 
  
acc_fc %>% autoplot(stroke_data, level=NULL) +
  theme_classic()

acc_fc %>% accuracy(stroke_data) %>% 
  dplyr::select(-c(.type, ME, MPE, ACF1)) %>%
  arrange(RMSE)

# Best fit model: need to generate the 95% PI)
acc_futures = acc_models %>%
  generate(h = "4 years", times = 100) %>%
  as_tibble() %>% group_by(Month, .model) %>%
  summarise(dist = distributional::dist_sample(list(.sim))) %>%
  ungroup() %>%
  as_fable(index = Month, key = .model,
           distribution = dist,
           response = "stroke_deaths_clean")

# Check the accuracy of the 95% predicted intervals (lower is better)
acc_futures %>% accuracy(stroke_data, 
                         measures = interval_accuracy_measures,
                         level = 95) %>% arrange(winkler)

# Extract Dec-2027 forecast at 80% from ETS
ets_dec_2028 = acc_fc %>%
  filter(.model == "ETS",
         Month == yearmonth("2028 Dec"))  %>%
  hilo(level=80) %>%
  unpack_hilo(`80%`) %>%
  rename(mean = .mean, LCI = `80%_lower`, UCI = `80%_upper`) %>%
  mutate(mean = round(mean),
         LCI = round(LCI),
         UCI = round(UCI)) 

deaths_forecasts = acc_futures %>% filter(.model == "ETS") %>%
  autoplot(stroke_data) + theme_bw() +
  labs(x = "Period",
       y = "Stroke deaths") +
    theme(legend.position = c(0.02, 0.96),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", 
                                     colour = "black")) +
  geom_vline(xintercept = yearmonth("2028 Dec"),
             linetype = "dashed", colour = "red3") +
  geom_hline(yintercept = ets_dec_2028$mean,
             linetype = "dashed", colour = "red3") +
  geom_point(data = ets_dec_2028,
             aes(x = Month, y = mean),
             size = 2, colour = "red3") +
  geom_text(data = ets_dec_2028,
            aes(x = Month, y = mean,
                label = paste0("(2028 Dec, ", mean, ")",
                "\n", "80% CI (",LCI, ", ", UCI,")")),
            vjust = -0.3, hjust = 1.05, size = 3, 
            fontface = "bold", colour = "red3")
deaths_forecasts

```


```{r}

stroke_data = stroke_data %>% 
  mutate(pop_million = population/1e6)

stroke_long = stroke_data %>%
  dplyr::select(Month, stroke_accidents_clean,
         stroke_cases_clean, stroke_deaths_clean) %>%
  pivot_longer(cols = -Month, names_to = "type", values_to = "count")

pop_max = max(stroke_data$pop_million)
count_max = max(stroke_long$count)
scale_factor = (pop_max / count_max)*4/5

ggplot() +
  geom_line(data = stroke_data,
            aes(x = Month, y = pop_million, colour = "pop_million"), size = 1.2) +
  geom_line(data = stroke_long, aes(x = Month,
                                    y = count * scale_factor,
                                    color = type), size = 1.2) +
  scale_y_continuous(name = "Population in millions",
    sec.axis = sec_axis(~ . / scale_factor,
                        name = "Stroke Events")) +
  scale_color_manual(values = c("pop_million" = "black", 
                            "stroke_accidents_clean" = "skyblue",
                            "stroke_cases_clean" = "coral",
                            "stroke_deaths_clean" = "red2"),
                     labels = c("Population", "Stroke admissions",
                                "Stroke Cases", "Stroke Deaths"),
                     name = "") + labs(x = "Period") + theme_minimal() + theme(legend.position = "bottom")
```


### Merged plots
```{r}

library(ggpubr)

ggarrange(cases_metrics, deaths_metrics,
          ncol=1, nrow=2)

ggarrange(cases_forecast, deaths_forecasts,
          ncol=1, nrow=2)

ggarrange(s3, s4, s5, s6, ncol=2, nrow=2)

top_row = ggarrange(s3, s5, ncol = 2)
middle_row = ggarrange(s4, ncol = 1)
bottom_row = ggarrange(s6, ncol = 1)
ggarrange(top_row, middle_row, bottom_row, ncol = 1,
                       heights = c(1, 1, 1))


```

